Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-10-03 manuscript.tex
organized folders, added temp

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, both positive and negative selection can remove genetically linked neutral variation, resulting in reduced $N_{e}$ and the level of polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male fitness can also reduce Y-linked $N_{e}$, making it difficult infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked polymorphism during the early stages of plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). Strikingly, we find that nucleotide diversity on the Y chromosome is 40 fold lower than on the X chromosome. We demonstrate that the magnitude of this reduction is not consistent with a purely neutral model based on the occurrence of female biased sex ratios in this species, or by reduced male $N_{e}$ caused by high variance in male fitness. Rather, using forward simulations, we show that the diversity reduction on the Y is consistent with the predicted effects of purifying selection removing deleterious mutations and linked neutral variation. We do not exclude the possibility that positive selection on beneficial mutations has also reduced Y chromosome variability, and suggest that both forms of linked selection have likely been important during plant sex chromosome evolution. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our results indicate that the low level of sequence diversity commonly observed on ancient Y chromosomes might evolve soon after sex chromosomes originate, with linked selection playing an important role during the early stages of their evolution.
\end{abstract}

>>>>

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, selection can remove genetically-linked neutral variation, resulting in reduced $N_{e}$ and the level of polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male fitness can also reduce Y-linked $N_{e}$, making it difficult infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked polymorphism during the early stages of plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). We find that nucleotide diversity on the Y chromosome is approximately 40 fold lower than on the X chromosome, suggesting an extensive and relatively rapid loss of diversity. We demonstrate that the magnitude of this diversit reduction is not consistent with a purely neutral model based on the occurrence of female biased sex ratios in this species, or by reduced male $N_{e}$ caused by high variance in male fitness. Rather, using forward simulations, we show that the diversity reduction on the Y is consistent with the removal of deleterious mutations and genetically-linked neutral variation. In particular, we find that .... hr realm...

Although we do not exclude the possibility that positive selection has also reduced Y chromosome variability, our simulations suggest that the reduction in diversity arising from purifying selection is sufficient to explain our observed patters of neutral variation. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our results suggest that the low level of sequence diversity commonly observed on ancient Y chromosomes might evolve relatively quickly after sex chromosomes originate, with selection against deleterious mutations playing an important role during the earliest stages of their evolution.
\end{abstract}
